---
title: "Stat 184"
author: "Brandon Montez"
output:
  pdf_document: default
  html_notebook: default
---

```{r}
library(laeken)
library(sdcMicro)
```

```{r}
require("laeken")
data("eusilc", package = "laeken")
```

# (a)
```{r}
sdc <- createSdcObj(dat = eusilc,
  keyVars = c("age", "pb220a", "pl030", "rb090", "hsize"),
  weightVar = "rb050",
  hhId = "db030")
```

# (b)
```{r}
risk <- get.sdcMicroObj(sdc, type="risk")$individual
hist(risk[, "risk"], main="Individual Risk", xlab="Individual Risk", col="blue")
```

It appears that most people have an individual risk that is small, but there is a cluster around 0.01 risk which is worth investigating. As there may be a high risk of re-identification.

## c

```{r}
hh_risk <- risk[, "hier_risk"]
hist(hh_risk, main="Household Risk", xlab="Household Risk", col="red")
```

If we consider 0.05 the threshold for too high of risk then the household risk evidently does exceed 0.05 for some observations. The household risk is generally below 20% but but there are some edge cases possibly suggesting non-uniformity in the distribution of key variables.

## d

```{r}
print(sdc, "risk")
```
The household risk is higher in general. We can see this from the histograms or from the information above. Given the hierarchical information, the expected number of re-identification is much higher. 

## e

```{r}
global_risk <- sum(risk[, "risk"])
global_risk
```

The estimated global risk using the simple summation of individual risk method says that we have around 20.94 expected individual re-identifications.

```{r}
set.seed(123)
subset_indices <- sample(nrow(eusilc), size = 0.1 * nrow(eusilc), replace = FALSE)
eusilc_subset <- eusilc[subset_indices, ]

sdc_subset <- createSdcObj(dat = eusilc_subset, 
  keyVars = c("age", "pb220a", "pl030", "rb090", "hsize"),
  weightVar = "rb050", 
  hhId = "db030")

risk_subset <- get.sdcMicroObj(sdc_subset, type="risk")$individual
global_risk_subset <- sum(risk_subset[, "risk"])

global_risk
global_risk_subset
```

It does appear that smaller and smaller subsets reduce the global risk. That is to be expected, as the global risk is a proportion of the number of individuals in the dataset. If we only have 1 individual, then our expected number of identifications could be at most 1. 

```{r}
print(sdc_subset, "risk")
```

This confirms our observation that the reduced subset has a lower global risk. The subset has redcued risk in both individual and household risk. 
