---
title: "Stat 184 HW 4"
author: "Brandon Montez"
output:
  pdf_document: default
  html_notebook: default
---
# Q1
## (a)
```{r}
library(synthpop)
data(SD2011)
```

## (b)
```{r}
library(diffpriv)
SD2011 <- SD2011[!is.na(SD2011$income),]

f <- function(X) mean(X)
M <- DPMechGaussian(target = f)

bs <- function(n, var=SD2011$income){var[sample.int(n=length(var),size=n, replace=TRUE)]}
M <- sensitivitySampler(M, oracle = bs, n = nrow(SD2011), m=10000)
sens_f <- M@sensitivity
```

## (c)

```{r}
sd_noise <- sens_f / 0.2

r <- function(x, M, mu=0.2){f(x) + rnorm(n=1, mean=0, sd = M@sensitivity / mu)}

replicate(5, r(SD2011$income, M))
```


## (d)

```{r}
SD2011 <- SD2011[SD2011$income > 0,]

f_log <- function(X) mean(log(X))
M_log <- DPMechGaussian(target = f_log) 
bs_log <- function(n, var=log(SD2011$income)){var[sample.int(n=length(var),size=n, replace=TRUE)]}
M_log <- sensitivitySampler(M_log, oracle = bs_log, n = nrow(SD2011), m=10000)
sens_f_log <- M_log@sensitivity

sd_noise_log <- sens_f_log / 0.2
r_log <- function(x, M, mu=0.2){f_log(x) + rnorm(n=1, mean=0, sd = M@sensitivity / mu)}

replicate(5, exp(r_log(SD2011$income, M_log)))
```
The log income model is better as it centers the data and more closely resembles a normal distribution. This makes sense as our sensitive variable is income, which typically has a skewed distribution.

# Q2
## (a)
```{r}
vars <- c("sex", "age", "placesize", "region", "edu", "socprof", "unempdur", "income", "marital")
SD2011_sub <- SD2011[,vars]

SD2011_syn <- syn(SD2011_sub, m=5)
```

## (b)

```{r}
compare.synds(SD2011_syn, SD2011_sub, vars = "income", stat = "counts", breaks = 25, table = TRUE)
```
The synthetic data appears to be extremely similar to the original data, given the very low S_pMSE value.

This is supported by visual inspection of the histograms. 